---
params:
  dataFile: ""
  line: ""
  desc: ""
title: "Mt Victoria Vermin Trappers Report"
subtitle: '`r paste0("Trap line: ", params$line)`'
author: "Report created by: Ben Anderson"
date: '`r paste0("Last run: ", Sys.Date())`'
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
  bookdown::pdf_document2:
    toc: true
    number_sections: true
    fig_width: 6
always_allow_html: true
urlcolor: blue
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)

library(dkUtils) # devtools::install_github("dataknut/dkUtils")

libsToLoad <- c("data.table",
                "flextable",
                "ggplot2",
                "lubridate")

dkUtils::loadLibraries(libsToLoad)

rmdParams <- list()

if(params$dataFile == ""){
  #notebook mode so params not set via YAML
  rmdParams$dataFile <- path.expand("~/Dropbox/Home/trapping/2024-06-02-telford.csv") # not open data
} else {
  rmdParams$dataFile <- params$dataFile
}
```

```{r loadData}

dt <- data.table::fread(rmdParams$dataFile)
dt[, date_nz := lubridate::ymd_hm(date, tz = "Pacific/Auckland")]
dt[, year := lubridate::year(date_nz)]
dataDate <- file.info(rmdParams$dataFile)[[4]]
dt[, supp_trap_id := `supp trap id`]
```

# Code

The code used to produce this report is [open source](https://github.com/dataknut/trap.nz) under an [Apache 2.0 license](https://github.com/dataknut/trap.nz?tab=Apache-2.0-1-ov-file#readme). Feel free to contribute. Or fork & go.

\newpage

# Introduction
This is a simple report of trapping activity and outcomes for the Mt Victoria Vermin Trappers' **`r params$line`** trap line `r params$desc`. It uses the latest 'complete trapline' records since `r min(as.Date(dt$date_nz))` downloaded from [trap.nz](https://trap.nz/node/169960) at `r dataDate`.

The **`r params$line`** trap line comprises `r uniqueN(dt$code)` traps made up of the types shown in Table \@ref(tab:trapTypes). This includes `r uniqueN(dt$supp_trap_id)` supplementary traps.

```{r trapTypes}
t <- dt[, .(N = uniqueN(code)), keyby = .(`trap type`)]

ft <- flextable::flextable(t)
autofit(set_caption(ft, caption = "Trap types"))
```


# Trap checking frequency

Figure \@ref(fig:checkDates) shows the frequency of records for each trap by month since the start of trapping. Normally we would hope there to be at least 2 checks per month. There are some clear gaps which will affect any subsequent 'trend' analysis.

```{r checkDates, fig.height=6, fig.cap="Trap checking frequency over time"}

dt[, month := floor_date(date_nz, "1 month")]

plotDT <- dt[, .(n = .N), keyby = .(Month = month, code)]
library(viridisLite)
ggplot2::ggplot(plotDT, aes(x = Month, y = code, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c() +
      theme(legend.position = "bottom",
          legend.direction = "horizontal") +
  labs(fill = "Number of records")
  
```


# Species caught

The species caught since the beginning of the monitoring dataset (`r min(as.Date(dt$date_nz))`) are shown in Figure \@ref(fig:species). Note that this does not (cannot) include any species killed by the A24 traps (or indeed any other traps) which have then been eaten as carrion or otherwise removed before the trap has been re-checked.

In addition the frequency of checking (see Figure \@ref(fig:checkDates)) is likely to affect the frequency of species capture since 'full' traps are unlikely to attract further kills until cleared.

```{r species, fig.cap="Species recorded"}

plotDT <- dt[, .(n = .N), keyby = .(Species = `species caught`)]

ggplot2::ggplot(plotDT[Species != "None" &
                         Species != "Unspecified"], aes(x = Species, y = n)) +
  geom_col() +
  coord_flip()

```

Figure \@ref(fig:speciesTimeline) shows the frequency of capture of different recorded species per month with additional smoothed fit lines for each species where sufficient data exists. 

This plot appears to be particularly affected by the frequency of checking patterns shown in Figure \@ref(fig:checkDates). However if we compare periods of relatively frequent checking it appears that the frequency of rat capture is lower in early 2024 than in (e.g.) early 2020. However it also appears that the frequency of mouse capture has increased.

```{r speciesTimeline, fig.cap="Frequency of species caught per month"}
plotDT <- dt[, .(n = .N), keyby = .(Species = `species caught`,
                                    Month = month)]

p <- ggplot2::ggplot(plotDT[Species != "None" &
                         Species != "Unspecified"], aes(x = as.Date(Month), y = n, colour = Species)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Month", y = "N caught per month") +
      theme(legend.position = "bottom",
          legend.direction = "horizontal",
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Date")

p + scale_x_date(date_labels="%b-%Y",date_breaks  ="3 month")

```

# Seasonal analysis

Do we find a monthly/seasonal pattern?

```{r speciesMonth, fig.cap="Species recorded by month"}
plotDT <- dt[, .(n = .N), keyby = .(Species = `species caught`,
                                    Month = lubridate::month(date_nz, label = TRUE))]

ggplot2::ggplot(plotDT[Species != "None" &
                         Species != "Unspecified"], aes(x = Month, y = n, colour = Species, group = Species)) +
  geom_point() +
  geom_smooth() +
  labs(caption = "All years")

```

# Summary

Goes here

